\bf{Neural Ordinary Differential Equations}

### 2025-03-20 paper.tex
some adjustments

\title{\bf{Neural Ordinary Differential Equations}}
\author{Prithvi Dake\\
Department of Chemical Engineering, University of California, Santa Barbara}
\date{\textit{ME255NN: Neural Networks for modeling, optimization and control} \\
\textit{Winter 2025}}

@@

@@

The NODE paper applies the same motivation to develop continuous normalizing flows
\citep{grathwohl:chen:bettencourt:sutskever:duvenaud:2018}.
Interested reader can refer the paper for proof of theorem on instantaneous change of variables
\citep{chen:rubanova:bettencourt:duvenaud:2018}.
The key idea is the neural ODE rhs is trained to conserve the probability density 
while maximizing log-likelihood to describe the data. The framework can be described 
as follows:
\begin{align}
    \begin{bmatrix}
        x_0 \\
        \log p(x) - \log p(x_0)
    \end{bmatrix}
    =
    \int_{t}^{0}
    \begin{bmatrix}
        f_{NN}(x, u; \theta) \\
        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
    \end{bmatrix} dt;
    \qquad
    \begin{bmatrix}
        x(t) \\
        \log p(x) - \log p(x_t)
    \end{bmatrix}   
    =
    \begin{bmatrix}
        x_t \\
        0
    \end{bmatrix}
\end{align}
We assume, $x_0 \sim \mathcal{N}(0, \sigma^2I)$ and thus the objective function is:
\begin{align}
    \min\limits_{\theta} V(x) = -\log p(x)
\end{align}
That's a powerful framework to model distributions (say a human face composed of pixels).
However, we are more interested in using similar generative framework for chemical kinetics. 
Thus, we again use the same example as shown in \cref{eq:atobtoc} but we also model 
how the uncertainty in the initial conditions decreases as we add more measurement points.
The results are shown in \cref{fig:cnf}. The system we solve is:

\begin{align}
    \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \\ \dot{\log p} \end{bmatrix}
    = \begin{bmatrix}
        -k_{1}c_A + k_{-1}c_B  \\
        2k_{1}c_A +  2k_{-1}c_B - k_2c_B \\
        k_{2}c_B\\
        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
    \end{bmatrix};
    &\qquad
    \begin{bmatrix} c_A(0) \\ c_B(0) \\ c_C(0) \\ \log p(0) \end{bmatrix} =
    \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
    \min\limits_{\theta} V(x) &= -\log p(x) + \norm{c - \tilde{c}}^2
\end{align}
The author would like to draw parallels to the classical Kalman filter \citep{kalman:1960} that also does the similar job,
but the continuous normalizing flow can be applied to any non-linear dynamical system.
\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth, page=3]{ABC_plot.pdf} 
    \caption{Results for chemical kinetics modeled as continuous normalizing
    flow. We start with a high uncertainty in the initial conditions. The spread in 
    probability decreases as we add more measurement points. Readers would like
    to draw similarities to Kalman filter, only in this case we can apply it to 
    any non-linear continuous dynamical system.}
    \label{fig:cnf}
\end{figure}

@@

\section{Data availability and Code}
All the figures in the review have been generated by the author. The code for the
figures can be found at

>>>>

\title{\bf{Neural Ordinary Differential Equations}}
\author{Prithvi Dake\\
Department of Chemical Engineering, University of California, Santa Barbara}
\date{\today\\
\textit{ME255NN: Neural Networks for modeling, optimization and control} \\
\textit{Winter 2025}}

@@

\noindent \textbf{Notation}: We keep our notation consistent with the one used in
\cite{ rawlings:mayne:diehl:2020}.

@@

The NODE paper applies the same motivation to develop continuous normalizing flows
\citep{grathwohl:chen:bettencourt:sutskever:duvenaud:2018}.
Interested reader can refer the paper for proof of theorem on instantaneous change of variables
\citep{chen:rubanova:bettencourt:duvenaud:2018}.
The key idea is the neural ODE rhs is trained to conserve the probability density 
while maximizing log-likelihood to describe the data. The framework can be described 
as follows:
\begin{align}
    \begin{bmatrix}
        x_0 \\
        \log p(x) - \log p(x_0)
    \end{bmatrix}
    =
    \int_{t}^{0}
    \begin{bmatrix}
        f_{NN}(x, u; \theta) \\
        -\mathrm{tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
    \end{bmatrix} dt;
    \qquad
    \begin{bmatrix}
        x(t) \\
        \log p(x) - \log p(x_t)
    \end{bmatrix}   
    =
    \begin{bmatrix}
        x_t \\
        0
    \end{bmatrix}
\end{align}
We assume, $x_0 \sim \mathcal{N}(0, \sigma^2I)$ and thus the objective function is:
\begin{align}
    \min\limits_{\theta} V(x) = -\log p(x)
\end{align}
That's a powerful framework to model distributions (say a human face composed of pixels).
However, we are more interested in using similar generative framework for chemical kinetics. 
Thus, we again use the same example as shown in \cref{eq:atobtoc} but we also model 
how the uncertainty in the initial conditions decreases as we add more measurement points.
The results are shown in \cref{fig:cnf}. The system we solve is:

\begin{align}
    \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \\ \dot{\log p} \end{bmatrix}
    = \begin{bmatrix}
        -k_{1}c_A + k_{-1}c_B  \\
        2k_{1}c_A +  2k_{-1}c_B - k_2c_B \\
        k_{2}c_B\\
        -\mathrm{tr} \left(\dfrac{\partial f_{NN}}{\partial c^T}\right)
    \end{bmatrix};
    &\qquad
    \begin{bmatrix} c_A(0) \\ c_B(0) \\ c_C(0) \\ \log p(0) \end{bmatrix} =
    \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
    \min\limits_{\theta} V(c) &= -\log p(c) + \norm{c - \tilde{c}}^2
\end{align}
The author would like to draw parallels to the classical Kalman filter \citep{kalman:1960} that also does the similar job,
but the continuous normalizing flow can be applied to any non-linear dynamical system.
\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth, page=3]{ABC_plot.pdf} 
    \caption{Results for chemical kinetics modeled as continuous normalizing
    flow. We start with a high uncertainty in the initial conditions. The spread in 
    probability decreases as we add more measurement points. Readers would like
    to draw similarities to Kalman filter, only in this case we can apply it to 
    any non-linear continuous dynamical system.}
    \label{fig:cnf}
\end{figure}

@@

\section{Data availability and Code}
All the figures in the review have been generated by the author. The code for the
figures can be found at \url{https://github.com/dakeprithvi/2025a_cnf}.
